---
title: "Gene Set Analysis Workflow Using gVenn"
subtitle: "Supplementary data S2"
output-file: "supp_data_S2"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#047857"
    theme: 
      - journal
      - custom.scss
    embed-resources: true
    fig-width: 7
    fig-height: 5
    fig-align: center
output-dir: docs
execute:
  echo: true
  warning: false
  message: false
---

# Add this copmutational notebook is available here: <link>

# Overview

This supplementary material demonstrates a complete gene set analysis workflow using the **gVenn** R package. The workflow includes computing overlaps between gene lists, visualizing results with Venn diagrams and UpSet plots, extracting overlap groups, and exporting results for downstream analysis.

# Installation

The gVenn package can be installed from Bioconductor or GitHub:

```{r}
#| label: installation
#| eval: false

# From Bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("gVenn")

# From GitHub (development version)
# install.packages("pak")
# pak::pak("ckntav/gVenn")
```

# Load Required Packages

```{r}
#| label: setup

library(gVenn)
library(knitr)
```

# Example Dataset

The **gVenn** package includes a synthetic gene list dataset (`gene_list`) comprising three sets of human gene symbols with designed overlaps. This dataset was generated from the first 250 gene symbols in the `org.Hs.eg.db` package using a reproducible random seed.

```{r}
#| label: load-data

# Load the example gene list dataset
data(gene_list)

# Display the structure of the dataset
str(gene_list)
```

The dataset contains three gene lists with the following sizes:

```{r}
#| label: dataset-summary

# Create a summary table
summary_df <- data.frame(
  "Gene List" = names(gene_list),
  "Number of Genes" = sapply(gene_list, length),
  check.names = FALSE
)

kable(summary_df, 
      caption = "Summary of example gene lists")
```

# Computing Overlaps

The `computeOverlaps()` function analyzes the intersection patterns across all gene lists and returns a structured object containing:

- A vector of all unique elements across sets
- A logical matrix indicating set membership for each element
- Category labels encoding the overlap pattern (e.g., "110", "101")

```{r}
#| label: compute-overlaps

# Compute overlaps between gene lists
gene_overlaps <- computeOverlaps(gene_list)

# Display the structure of the result
class(gene_overlaps)
names(gene_overlaps)
```

## Overlap Matrix

The overlap matrix shows which genes belong to which sets:

```{r}
#| label: overlap-matrix

# Display the first 10 rows of the overlap matrix
kable(head(gene_overlaps$overlap_matrix, 10),
      caption = "First 10 rows of the overlap matrix (TRUE = gene present in set)")
```

## Overlap Categories

Each gene is assigned a category code representing its overlap pattern:

```{r}
#| label: overlap-categories

# Show the distribution of overlap categories
category_table <- table(gene_overlaps$intersect_category)
category_df <- data.frame(
  "Overlap Pattern" = names(category_table),
  "Number of Genes" = as.integer(category_table),
  check.names = FALSE
)

kable(category_df,
      caption = "Distribution of overlap patterns")
```

**Interpretation of overlap patterns:**

- `100`: Genes only in set A (random_genes_A)
- `010`: Genes only in set B (random_genes_B)
- `001`: Genes only in set C (random_genes_C)
- `110`: Genes in A ∩ B (not in C)
- `101`: Genes in A ∩ C (not in B)
- `011`: Genes in B ∩ C (not in A)
- `111`: Genes in A ∩ B ∩ C (all three sets)

# Visualization

## Venn Diagram

The `plotVenn()` function creates area-proportional Venn diagrams to visualize the overlaps:

```{r}
#| label: fig-venn-basic
#| fig-cap: "Basic Venn diagram showing overlaps between three gene lists"

# Create a basic Venn diagram
plotVenn(gene_overlaps)
```

### Customized Venn Diagram

The appearance can be customized by adjusting colors, transparency, labels, and other parameters:

```{r}
#| label: fig-venn-custom
#| fig-cap: "Customized Venn diagram with adjusted colors and transparency"

# Create a customized Venn diagram
plotVenn(gene_overlaps,
         fills = list(fill = c("#2B70AB", "#FFB027", "#3EA742"), alpha = 0.6),
         edges = list(col = "gray30", lwd = 1.5),
         labels = list(col = "black", fontsize = 12, font = 2),
         quantities = list(type = c("counts", "percent"), 
                          col = "black", fontsize = 10),
         main = list(label = "Gene Set Overlaps", 
                    fontsize = 14, font = 2, col = "navy"),
         legend = list(side = "right", fontsize = 10))
```

## UpSet Plot

For larger numbers of sets (>3), UpSet plots provide a clearer alternative to Venn diagrams:

```{r}
#| label: fig-upset
#| fig-cap: "UpSet plot showing intersection sizes between gene lists"
#| fig-width: 7
#| fig-height: 4

# Create an UpSet plot
plotUpSet(gene_overlaps)
```

### Customized UpSet Plot

The UpSet plot can also be customized with colors:

```{r}
#| label: fig-upset-custom
#| fig-cap: "Customized UpSet plot with colored combination matrix"
#| fig-width: 7
#| fig-height: 4

# Create a customized UpSet plot with colored dots
plotUpSet(gene_overlaps, 
          comb_col = c("#2B70AB", "#FFB027", "#3EA742", "#CD3301", 
                      "#9370DB", "#008B8B", "#D87093"))
```

# Extracting Overlap Groups

The `extractOverlaps()` function separates genes into distinct groups based on their overlap patterns:

```{r}
#| label: extract-overlaps

# Extract genes grouped by overlap pattern
gene_groups <- extractOverlaps(gene_overlaps)

# Display the number of genes per group
group_sizes <- sapply(gene_groups, length)
group_df <- data.frame(
  "Overlap Group" = names(group_sizes),
  "Number of Genes" = as.integer(group_sizes),
  check.names = FALSE
)

kable(group_df,
      caption = "Number of genes in each overlap group")
```

## Examining Specific Groups

Individual overlap groups can be accessed for downstream analysis:

```{r}
#| label: specific-groups

# Extract genes present in all three sets (group_111)
genes_in_all_three <- gene_groups[["group_111"]]

cat("Genes present in all three sets (A ∩ B ∩ C):\n")
print(genes_in_all_three)

# Extract genes unique to random_genes_A (group_100)
genes_unique_to_A <- gene_groups[["group_100"]]

cat("\n\nFirst 20 genes unique to random_genes_A:\n")
print(head(genes_unique_to_A, 20))
```

## Summary Table of All Groups

```{r}
#| label: all-groups-table

# Create a detailed table showing first 5 genes from each group
group_preview <- lapply(names(gene_groups), function(group_name) {
  genes <- gene_groups[[group_name]]
  preview <- if(length(genes) <= 5) {
    paste(genes, collapse = ", ")
  } else {
    paste(c(head(genes, 5), "..."), collapse = ", ")
  }
  data.frame(
    Group = group_name,
    Count = length(genes),
    "Example Genes" = preview,
    check.names = FALSE
  )
})

group_preview_df <- do.call(rbind, group_preview)

kable(group_preview_df,
      caption = "Summary of all overlap groups with example genes")
```

# Exporting Results

## Export to Excel

The `exportOverlaps()` function exports each overlap group to a separate sheet in an Excel file:

```{r}
#| label: export-excel
#| eval: false

# Export overlap groups to Excel
exportOverlaps(gene_groups,
               output_dir = "results",
               output_file = "gene_overlap_groups",
               with_date = TRUE,
               verbose = TRUE)
```

This creates an Excel file with one sheet per overlap group, making it easy to:

- Review genes in each category
- Perform functional enrichment analysis
- Share results with collaborators
- Import into other analysis tools

## Saving Visualizations

Visualizations can be exported in multiple formats (PDF, PNG, SVG):

```{r}
#| label: save-viz
#| eval: false

# Create a Venn diagram
venn_plot <- plotVenn(gene_overlaps)

# Save as PDF
saveViz(venn_plot,
        output_dir = "figures",
        output_file = "gene_venn_diagram",
        format = "pdf",
        width = 6,
        height = 4)

# Save as high-resolution PNG
saveViz(venn_plot,
        output_dir = "figures",
        output_file = "gene_venn_diagram",
        format = "png",
        width = 6,
        height = 4,
        resolution = 300)

# Save with transparent background for presentations
saveViz(venn_plot,
        output_dir = "figures",
        output_file = "gene_venn_diagram_transparent",
        format = "png",
        bg = "transparent")
```

# Downstream Analysis Applications

The extracted overlap groups can be used for various downstream analyses:

## Functional Enrichment Analysis

```{r}
#| label: enrichment-example
#| eval: false

# Example: Perform GO enrichment on genes unique to set A
library(clusterProfiler)
library(org.Hs.eg.db)

genes_unique_A <- gene_groups[["group_100"]]

# Convert gene symbols to Entrez IDs
entrez_ids <- bitr(genes_unique_A, 
                   fromType = "SYMBOL",
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

# Perform GO enrichment
go_results <- enrichGO(gene = entrez_ids$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       ont = "BP",
                       pvalueCutoff = 0.05)
```

## Comparing Shared vs. Unique Genes

```{r}
#| label: comparison-table

# Compare genes shared across all sets vs. unique to each set
comparison_df <- data.frame(
  Category = c("Shared (all 3 sets)",
               "Unique to A",
               "Unique to B", 
               "Unique to C"),
  Count = c(
    length(gene_groups[["group_111"]]),
    length(gene_groups[["group_100"]]),
    length(gene_groups[["group_010"]]),
    length(gene_groups[["group_001"]])
  ),
  Percentage = c(
    round(100 * length(gene_groups[["group_111"]]) / length(gene_overlaps$unique_elements), 1),
    round(100 * length(gene_groups[["group_100"]]) / length(gene_overlaps$unique_elements), 1),
    round(100 * length(gene_groups[["group_010"]]) / length(gene_overlaps$unique_elements), 1),
    round(100 * length(gene_groups[["group_001"]]) / length(gene_overlaps$unique_elements), 1)
  )
)

kable(comparison_df,
      caption = "Comparison of shared vs. unique genes")
```

# Working with Your Own Data

To use **gVenn** with your own gene lists:

```{r}
#| label: custom-data-example
#| eval: false

# Example: Creating gene lists from your own data
my_gene_lists <- list(
  Treatment_A = c("TP53", "BRCA1", "EGFR", "MYC", "KRAS"),
  Treatment_B = c("TP53", "BRCA1", "PTEN", "AKT1", "PIK3CA"),
  Control = c("EGFR", "MYC", "AKT1", "BRAF", "NRAS")
)

# Compute overlaps
my_overlaps <- computeOverlaps(my_gene_lists)

# Visualize
plotVenn(my_overlaps)

# Extract groups
my_groups <- extractOverlaps(my_overlaps)

# Export results
exportOverlaps(my_groups,
               output_dir = "my_results",
               output_file = "my_gene_overlaps")
```

# Session Information

```{r}
#| label: session-info

sessionInfo()
```

# References

1. **gVenn Package**  
   Tav, C. (2025). gVenn: Proportional Venn and UpSet Diagrams for Gene Sets and Genomic Regions. 
   R package. https://github.com/ckntav/gVenn

2. **eulerr Package**  
   Larsson, J. (2023). eulerr: Area-Proportional Euler and Venn Diagrams with Ellipses. 
   https://CRAN.R-project.org/package=eulerr

3. **ComplexHeatmap**  
   Gu, Z., Eils, R., & Schlesner, M. (2016). Complex heatmaps reveal patterns and correlations 
   in multidimensional genomic data. *Bioinformatics*, 32(18), 2847-2849. 
   doi:10.1093/bioinformatics/btw313

4. **Gene Set Analysis**  
   Subramanian, A., et al. (2005). Gene set enrichment analysis: a knowledge-based approach 
   for interpreting genome-wide expression profiles. *PNAS*, 102(43), 15545-15550.
